### Problem 2

Read [this student project](https://home.csulb.edu/~jchang9/m247/m247_fa11_David_Diego_Alissa_Daniel.pdf) about linear algebra and ciphers, from [Prof. Jen-Mei Chang](https://home.csulb.edu/~jchang9/index.html)'s Linear Algebra course at California State University, Long Beach.

For convenience, I have written a function that converts a string of text (letters a through z, ignoring case, as well as space, comma, and period) into a matrix with entries 1 - 29, where the letters are 1 - 26, the space is 27, and the comma is 28, and the period is 29. We'll work with text in groups of N characters, so the matrix produced will have N rows. There's no actual ciphering going on yet. We're just implementing a way of representing letters (and space and period) straightforwardly as numbers. 
```{r}
texttomatrix <- function(text,N){
  key <- c(LETTERS," ",",",".")
  text <- toupper(text)
  newlength <- ceiling(nchar(text)/N)*N
  text <- str_pad(text,newlength,"right")
  tmp <- unlist(strsplit(text, split = ""))
  result <- matrix(match(tmp, key), nrow = N)
  return(result)
}
```
For example, using blocks of 20 letters,
```{r}
M <- texttomatrix("Once upon a time there was a very cute puppy.",20)
M
```

I have also written a function that takes this type of matrix and turns it back into text.
```{r}
matrixtotext <- function(M){
  key <- c(LETTERS," ",",",".")
  M <- as.numeric(M)
  result <- trimws(paste(key[M], collapse = ""))
  return(result)
}
```
For example,
```{r}
matrixtotext(M)
```

a. Experiment with each line of code in the `texttomatrix` and `matrixtotext` functions I wrote for you. Explain what the built-in R functions `LETTER`, `toupper`, `ceiling`, `nchar`, `str_pad`, `strsplit`, `unlist`, `match` `as.numeric` (applied to something that is already a numerical matrix), `paste` (with `collapse = ""`), and `trimws` do in their contexts above.

b. In the example usage of `texttomatrix` above, why are there repeated values of `27` at the end?

c. Suppose you are receiving continuous transmissions of ciphertext (as numbers, stored in a vector) one sentence at a time, and you want to decipher them real time as they come in. I've created a 20 x 20 cipher matrix for you below.
```{r}
set.seed(123)
N <- 20
c <- sample(-300:300, N*N, replace = TRUE)
C <- matrix(c, nrow = N)
```
What is a computationally efficient way to decode the incoming text? (Hint: in this context, "computationally efficient" means that you should do as few matrix inversions as possible.) Implement your method on the four encoded sentences below.
```{r}
s1 <- c(9450, 5309, 3621, 7196, -2988, -6711, -3540, -12496, -5157, 
3091, 11463, -6858, -5570, -12771, 1393, 2917, 11056, -20835, 
-483, -6306, 19831, -2548, 3807, 2942, 14025, -17369, 8327, -18494, 
1261, 587, 5851, 1989, 12463, -17387, -371, 14754, -18495, -21795, 
4221, -7976, 22298, -1155, -889, 9549, -7929, -19996, 596, -15147, 
-7841, 9887, 9008, -9997, 5413, -20790, 11273, 36732, -9755, 
-25491, 3994, -3112, 28548, -4384, 15685, 11222, 602, -13878, 
-18383, -13174, 6597, -2755, 8172, -877, 4499, -25620, -4261, 
20467, -219, -22349, 2888, -17339, 11282, -6043, -1484, -15430, 
-4917, -13373, -10290, -8215, -8048, 2777, 6544, -6741, 2446, 
-22471, -1771, 7768, 7386, -15946, -3617, -9760, 20418, -3641, 
5066, 1191, 11842, -10134, -1170, -15963, -6142, 9531, 14026, 
-5745, 6187, -22898, 4486, 16147, -12695, -27109, -7824, -4971, 
19333, 7537, 4819, 11934, 3938, -22670, -15963, -7744, 930, 1916, 
15038, 3286, -1167, -33460, 13951, 16321, -554, -22727, -6380, 
-19020)
s2 <- c(12257, -54, 4102, 10792, 13035, -12547, 6602, -16035, 1342, 
-2884, 10784, 8201, 12006, -19133, 8275, 16190, -3014, -16689, 
-10090, -19441, 17112, -6130, 7855, -2562, 3886, -9700, 879, 
-11750, -5939, -8150, 16510, 7626, 984, -26116, -1671, 23490, 
-8749, -13703, -1382, -14416, 17760, 3999, -8006, 6444, -2585, 
-30757, -9462, -14277, -10714, -3669, 12041, -9829, 14059, -24666, 
10985, 15850, 1997, -31693, 6542, -16217)
s3 <- c(26646, -3881, 12186, 5296, 5851, -15292, -5619, -16723, 4438, 
5286, 8983, 2228, 2777, -22077, 3172, 9966, -2785, -21675, 1202, 
-23630, 12892, 12892, 7848, 22079, 5985, -21387, -7324, -14668, 
5065, -12738, 11710, 10289, 6385, -26435, 2671, 4816, -727, -19629, 
94, -12211, 17866, 14765, 1712, 14937, -3829, -18954, -8066, 
-16352, -2792, 215, 12812, -7781, -986, -20588, 4995, 2596, 14428, 
-24824, 2490, -9633, 17008, -13546, 11303, -2742, 2928, -11146, 
-7879, -14162, 882, -13636, 3616, 8864, 4177, -24202, -12623, 
24262, -12028, -16351, -8235, -12171, 24614, 4388, 5792, 22644, 
11521, -17583, -10650, -10978, 6748, -1825, 10898, 2814, 17003, 
-28130, 12455, 22251, -8591, -18133, -5319, -13905, 23691, 6196, 
7317, 10076, 7786, -25016, 391, -14459, 9001, -1869, 3784, 7241, 
-1578, -22724, -582, 11371, -8364, -21870, 6535, -22048, 20794, 
-9154, 10528, -5460, -1530, -2975, -5549, -14397, 45, 7178, 6229, 
1101, 576, -25505, -3055, 13523, -9350, -20866, -5945, -8217, 
14687, 1212, 12545, 1764, -2029, -10481, -6468, -13051, 17528, 
2402, -1993, 8729, -1984, -18144, -9916, -1302, -59, -19224, 
246, -16874, 28697, -2781, 2653, 473, 5410, -10567, -7507, -13103, 
-9947, 377, 19249, -4524, 10831, -20558, 12970, 22566, 460, -15430, 
406, -13638, 14711, 15673, 6062, 11700, -2128, -20559, -8194, 
-9284, -3626, -10374, 18417, 416, -5122, -21770, -6145, 6060, 
2635, -20511, 11185, -8686, 25228, -802, -9449, -2936, 4078, 
-22791, -4729, -11215, -8228, 1188, 10163, -602, 7382, -35695, 
9835, 24156, -12914, -31081, -5853, -17573)
s4 <- c(18390, -9213, 10878, 3440, 14930, -14154, 1349, -16994, 5298, 
-14517, 661, 19714, 12753, -25238, -8494, 13875, -24909, -17677, 
-4572, -18219, 17829, 2751, 5151, 4769, -11285, -26131, -9831, 
-15433, 6015, -6129, -1114, 369, 834, -22692, -9512, 13806, -3149, 
-22684, 11418, -7372, 14425, 559, 8971, -12295, -9532, -19671, 
-6072, -12348, -4532, 6398, 11662, 91, -8621, -20767, -579, 4304, 
59, -13988, 7671, -7998, 24230, 2237, 9027, 14894, 773, -7496, 
1796, -19271, 5465, 7939, 7997, 8534, 4050, -27248, 9452, 18019, 
-9729, -25964, -5969, -13151, 22985, -3064, -3907, -7897, 302, 
-9648, 1379, -7368, 3609, 3785, 668, -537, 2458, -24440, 4419, 
27129, -4314, -15508, 788, -16127, 20777, -997, 9766, 6917, 1948, 
-19196, -6312, -14687, -2036, 1444, 21437, -3017, 6301, -29331, 
2672, 29133, -5587, -20334, 690, -6530, 38326, -1646, -3518, 
3654, 8791, -21585, -7853, -15641, -4588, -9464, 6570, -1854, 
15369, -26819, 6330, 25957, -6318, -22058, 3306, -21971, 20162, 
5481, 9722, 22192, 2231, -12427, -8038, -11100, 12957, -1821, 
9125, 96, 8344, -18131, -6407, 18117, -3965, -20393, 4167, -6620, 
16811, -238, 92, 1658, 6248, -26653, 6911, -15480, 4910, 2062, 
663, 4928, 11082, -24141, 2983, 20715, -15901, -18748, 6241, 
-9508, 16391, 2099, 90, -250, -6964, -29248, -7943, -10799, 3087, 
9031, 2863, 5140, -4085, -29106, 13656, 14449, -5643, -20804, 
-1661, -14038)
```
