---
title: "Activity: R Bootcamp"
author: "Insert group members' full names here"
output: html_document
---

### Objectives

* Create your first in-class R Markdown
* Create matrices in R
* Evaluate polynomials in R

### Procedure

* Please collaborate on this activity with one person as scribe (typist). There should only be one laptop open. We're aiming for actual collaboration here, not two or three people working in parallel.

* It doesn't matter if you finish the whole activity during class.

* It **does** mater if you focus on helping each other learn. Helping each other learn is the point of this activity. Completing the problems is merely a mechanism to guide you towards learning.

* It **does** mater if everyone participates. The expectation is that every group member participates, where the definition of "participation" can include asking questions, answering questions, brainstorming, or any other form of meaningful engagement.

* In the last three minutes of class, before you leave, please knit your .Rmd file to html, print the html to .pdf, and have the scribe submit it as a group assignment through GLOW.

* Don't forget that the solutions are available on GLOW.

### Problem 0

Recalling that academic integrity policies apply even in an ungraded course, please disclose any irregular circumstances related to participation in your group (for instance, a group member was missing and did not contribute to this document).

### Problem 0 Solution

Your answer goes here.

### Problem 1

Generate the following matrices with as little code as possible.

a. 

$$\begin{bmatrix}
1 & 1 & 1 & 1 & 1 & 1 & 1 \\
1 & 1 & 1 & 1 & 1 & 1 & 1 \\
1 & 1 & 1 & 1 & 1 & 1 & 1 \\
1 & 1 & 1 & 1 & 1 & 1 & 1 \\
1 & 1 & 1 & 1 & 1 & 1 & 1 \\
1 & 1 & 1 & 5 & 5 & 5 & 1 \\
1 & 1 & 1 & 5 & 5 & 5 & 1 \\
1 & 1 & 1 & 1 & 1 & 1 & 1 
\end{bmatrix}$$

b.

$$\begin{bmatrix}
1 & 2 & 3 & \cdots & 10 \\
2 & 4 & 6 & \cdots & 20 \\
3 & 6 & 9 & \cdots & 30 \\
\vdots & \vdots & \vdots & \ddots & \vdots \\
8 & 16 & 24 & \cdots & 80
\end{bmatrix}
$$

c.

$$\begin{bmatrix}
1 & 1 & 1 & 1 & 1 \\
4 & 4 & 4 & 4 & 4 \\
9 & 9 & 9 & 9 & 9 \\
\vdots & \vdots & \vdots & \vdots & \vdots \\
121 & 121 & 121 & 121 & 121
\end{bmatrix}
$$

### Problem 1 Solution

a. Your solution goes here.

b. Your solution goes here.

c. Your solution goes here.

### Problem 2 

Here is an example of how to create a function in R:
```{r}
P <- function(x){
  y <- 2*x^5 - 3*x^4 + x^3 - 10*x^2 + 1
  return(y)
}
```
I can evaluate the function at a specific value of $x$:
```{r}
P(2)
```
or at multiple values of $x$:
```{r}
xx <- 3:7
P(xx)
```

a. Write a function to evaluate the same polynomial, $P(x)=2x^5-3x^4+x^3-10x^2+1$, but this time, let the inputs to the function be a vector containing the coefficients of $P$ and then a vector of values of $x$ at which you want to evaluate the polynomial. It should look like
```{r eval = FALSE}
P <- function(a,x) {
  
  # Your code goes here
  
}
```
where I have used $a$ to represent the vector of coefficients in order of highest degree to lowest degree (making sure to account for terms that have coefficients of zero). The point of doing this is that now don't need to change our function each time we want to evaluate a different polynomial. We can just pass in a different set of coefficients. **Before you start coding**, make a plan (in words) for how to do the calculation. Report your plan, write the code, use it to evaluate the polynomial at $x = 3, 4, 5, 6, 7$, and verify that the answers match what we computed above

Hints:

* Actually create a plan first. The real intellectual challenge of the problem is figuring out how to do the calculation in a nice, tight, efficient way. Turning the idea into code is only the second most important part.
* Avoid using a for-loop at all costs.
* The command `outer` with `FUN = "^"` will be helpful.
* My function to evaluate $P(x)$ was under 10 lines of brief code (not including comments).

### Problem 2 Solution

Your solution goes here.

### Key Commands

* `matrix` to create a matrix
* `:` (colon operator) to create a vector of integers
* `function` to define a function
* `length` to get length of a vector
* `rep` to copy a number/vector multiple times
* `t` to take transpose of a matrix or vector
* `%o%` to take outer product of two vectors
* `outer` to take outer product of two vectors (more flexible command)
* `rowSums` to get the sum of each row of a matrix