---
title: "Activity - Fundamentals of Linear Systems"
output:
  rmarkdown::html_document:
  header-includes:
   - \usepackage{amsmath}
---

\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\mat}[1]{\mathbf{#1}}

Load necessary packages:

```{r}
library(tictoc)
```

**Remember to show your work / give your reasoning, as applicable.**

### Problem 1

Which of the following sets of vectors are bases for $\mathbb{R}^2$?

-   $\{(0, 1), (1, 1)\}$
-   $\{(1, 0), (0, 1), (1, 1)\}$
-   $\{(1, 0), (-1, 0)\}$
-   $\{(1, 1), (1, -1)\}$
-   $\{(1, 1), (2, 2)\}$
-   $\{(1, 2)\}$

### Problem 1 solution

Your solution goes here.

### Problem 2

What is the dimension of the intersection of the following two planes in $\mathbb{R}^3$?

$$
x + 2y - z = 0, \ \ \ 3x - 3y + z = 0
$$

### Problem 2 Solution

Your solution goes here.

### Problem 3

Solve (by hand) the system below.

$$
\begin{pmatrix}1 & 2 & 0 \\ 3 & 2 & 4 \\ -2 & 1 & -2 \end{pmatrix} \begin{pmatrix} x \\ y \\ z \end{pmatrix} = \begin{pmatrix} 1 \\ 7\\ -1 \end{pmatrix}
$$

### Problem 3 Solution

Your solution goes here.

### Problem 4

Put the augmented coefficient matrix for the system of equations

$$
\begin{aligned}
x+y+z&=2\\
x+3y+3z&=0\\
x+3y+6z&=3
\end{aligned}
$$

into row echelon form.

### Problem 4 Solution

Your solution goes here.

### Problem 5

I've written a function called `eliminate` that performs Gaussian elimination in order to transform a matrix to row echelon form.

```{r}
eliminate <- function(A, tol = 10^-8) {
  n <- nrow(A)
  for ( j in 1:(n-1) ) {
    pivot <- A[j,j]
    if (abs(pivot) < tol) stop('zero pivot encountered')
    for ( i in (j+1):n ) {
      A[i,] <- A[i,] - A[i,j]/pivot * A[j,]
    }
  }
  return(A)
}
```

Demonstrate $\mathcal{O}(n^3)$ complexity by doing the following. Run the `eliminate` command on an $n \times n$ random matrix (generated using the `runif` command) for each of $n=270,\ 540,\ 810,\ 1080,\ 1350$, saving the run time for each value of $n$. To save the run time, you can use something like

```{r eval = FALSE}
tic()
2+2 # Just as an example
T1 <- toc()
t1 <- T1$toc - T1$tic
```

Note: the computations will not be instantaneous, and you may have to wait several minutes for them to finish. I urge you to use the `cache = TRUE` option in your code block, which will prevent R from re-running the block each time you knit, so long as the block hasn't changed. Instead, R will execute the code on the first knit, but it will save the answer and just re-use that answer the next time the code is knit. You should keep this option in mind as a possibility for all future coding you do in R markdown.

Additionally, you should "seed" the random number generator. This means that you'll produce the same random matrices every time, which will allow the result of the code block to remain the same, which will allow successful caching. After the three tick marks that start a code block, you'll write `{r cache = TRUE}`. Then your next line will seed the random number generator with `set.seed(123)` where the number 123 is totally arbitrary. You can use any number you like; just don't change it after you choose it.

Finally, use the `lm` command to fit a line to the points $(\log n,\log t)$ and find the slope.

### Problem 5 solution

Your solution goes here.

### Problem 6

Compute by hand the eigenvalues of the matrices

$$
\mat{A}=\begin{pmatrix} 1 & 1000 \\ 0 & 1\end{pmatrix}, \qquad \widetilde{\mat{A}}=\begin{pmatrix} 1 & 1000 \\ 0.001 & 1\end{pmatrix}.
$$

Would you say that the problem of computing eigenvalues is well-conditioned or ill-conditioned? Check the condition number of $\mat{A}$ using the `kappa` command with the option `exact = TRUE`. For concreteness, use the 2-norm, though there is nothing special about that choice.

### Problem 6 Solution

Your solution goes here.