---
title: "Pset - Principal Component Analysis"
output:
  rmarkdown::html_document:
  header-includes:
   - \usepackage{amsmath}
---

Load necessary packages:

```{r, message = FALSE}
library(Matrix)
library(imager)
library(ggplot2)
```

### Problem 1

By completing this problem, you'll learn how PCA can be used to compress images. Let's download an image of a famous meme.

```{r cache = TRUE}
url <- "http://www.slate.com/content/dam/slate/articles/news_and_politics/the_slate_quiz/authorPortraits/pronounce_doge4.jpg.CROP.promo-xlarge2.jpg"
doge <- load.image(url)
```

It's a bit more straightforward to work with black and white images, so let's convert this color image to black and white and display it.

```{r cache = TRUE}
dogebw <- grayscale(doge)
plot(dogebw, axes=FALSE)
```

One slightly confusing thing about images in `R` is that for a matrix that stores an image, the first index counts columns and the second index counts rows. Why? I have no idea. But for example, the code below will plot the first 400 columns of the image. To extract those columns, we'll need to convert the image to regular matrix, then do the extraction, then covert back to an image before plotting.

```{r cache = TRUE}
plot(as.cimg(as.matrix(dogebw)[1:400,]), axes = FALSE)
```

a\. The image is represented as a matrix of numbers, with each number representing the greyscale value of the pixel. Therefore, it takes as many numbers as there are pixels in the image to specify the image. How many pixels are there in the image?

b\. Perform PCA on the image by treating each column of the image (meaning each row of the matrix) as an observation. Then, reconstruct the image obtained by keeping 1, 3, 10, and 30 principal components and plot each reconstruction.

One technical hint: when we do PCA, we want our data centered around the origin, and `R` does this by default. After we reconstruct an image from principal components, we have to un-do that centering. For instance, suppose we wanted to reconstruct the original image, and suppose we saved the output of the `prcomp` command in a variable called `pca`. Then we would run something like

```{r eval = FALSE}
recon <- pca$x %*% t(pca$rotation)
recon <- scale(recon, center = -1*pca$center, scale = FALSE)
```

c\. How many components do you need to capture 99.9% of the variation in the data? Plot the reconstruction using that many components.

d\. How many numbers do you need to represent this reconstruction. Note: this is NOT the number of pixels in the image, which hasn't changed. Why? Because you made the image from a product of matrices. So you only need to keep track of however many numbers there are in the first matrix plus however many numbers there are in the second matrix.

e\. Divide your answer from (d) by your answer from (a) and comment on the level of image compression achieved.

### Problem 1 Solution

a\. Your solution goes here.

b\. Your solution goes here.

c\. Your solution goes here.

d\. Your solution goes here.

e\. Your solution goes here.