---
title: "Pset - Eigenvalues I"
output:
  rmarkdown::html_document:
  header-includes:
   - \usepackage{amsmath}
---

Load necessary packages:

```{r, message = FALSE}
library(pracma)
library(igraph)
```

### Problem 1

In this problem, you'll work with the famous Google PageRank algorithm. PageRank is sometimes known as the world's largest power iteration.

Let's generate a (random) network of 7 websites. The adjacency matrix is stored in $\textbf{A}$ and I have also visualized the network for you.

```{r cache = TRUE}
set.seed(123)
n <- 7
B <- matrix(rbinom(n^2, size = 1, prob = 0.3), nrow = n)
A <- B - diag(diag(B))
G <- graph_from_adjacency_matrix(A)
plot(G)
```

The purpose of this exercise is to understand how PageRank works, and what it has to do with power iteration. Imagine that you initially assign each node an equal PageRank score (such that the scores sum to one). We store these guesses in a vector $\mathbf{x}$. Then, as described by Jon Kleinberg: "Each node divides its current PageRank equally across its out-going links, and passes these equal shares to the nodes it points to. Each node updates its new PageRank to be the sum of the shares it receives."

One then repeats this process over and over until the PageRank scores converge, and those converged values are the final importance ratings of the nodes.

a\. Construct the matrix $\mathbf{M}$ which, when applied to $\mathbf{x}$, represents one iteration of PageRank. Hint: the matrix $\mathbf{M}$ we're after is not the matrix $\mathbf{A}$, though it certainly depends on $\mathbf{A}$. Another hint: the way to figure this out is to grab a pencil and paper and start doing the PageRank iteration by hand so you can see exactly what is involved in terms of matrix computation.

b\. Now you have the matrix $\mathbf{M}$ that does one iteration of PageRank when applied to $\mathbf{x}$. Since you want to do many iterations (until convergence), you are repeatedly multiplying $\mathbf{x}$ by $\mathbf{M}$. This is the definition of power iteration, and if it converges, it will converge to the dominant eigenvector of $\mathbf{M}$.

Use power iteration to find the PageRank values for each website. Stop your iteration when the 2-norm of successive approximations is less that $5 \times 10^{-10}$. After each step of power iteration, make sure to normalize $\mathbf{x}$ so that its entries sum to one. When you are done iterating, report which web site has the highest PageRank. Give the number of the node (1 through 7) and also state its actual PageRank score.

### Problem 1 Solution

a\. Your solution goes here.

b\. Your solution goes here.