---
title: "Pset - R Bootcamp"
output:
  rmarkdown::html_document:
  header-includes:
   - \usepackage{amsmath}
---

Load necessary packages.

```{r message = FALSE}
library(tictoc)
```

### Problem 1

In the 1999 movie Office Space, a character creates a program that takes fractions of cents that are truncated in a bank's transactions and deposits them into his own account. This is not a new idea, and hackers who have actually attempted it have been arrested. In this exercise you will explore this scam. The point of this problem is to practice some R skills.

Assume the following details:

* You have access to 10,000 bank accounts.

* Initially, the bank accounts have balances that are uniformly distributed between $100 and $100,000.

* The annual interest rate on the accounts is 5%.

* Interest is compounded each day and added to the accounts, except that fractions of a cent are truncated.

* The daily interest rate is thus $.05/365$.

* The truncated fractions are deposited into an illegal account that initially has a balance of $0.

* The illegal account can hold fractional values and it also accrues daily interest.

* Assume that each day, interest is earned (on the main account and the illegal account) first, and then skimming from the main account into the illegal account happens.

Your job is to write an R script that simulates this situation and finds how long it takes for the illegal account to reach a million dollars.

Here is some R help.

The following code generates the initial accounts:
```{r eval = FALSE}
accounts <- runif(10000,100,100000)
accounts <- floor(accounts*100)/100
```
The first line sets up 10,000 accounts with values uniformly between 100 and 100,000. The second line removes the fractions of cents (look at the data before and after that line is applied).

For the main part of your code, I strongly recommend a `while` loop, which would look something like:
```{r eval = FALSE}
while (illegal < 1000000) {
  do stuff here
}
```

### Problem 1 Solution

Your solution goes here.

### Problem 2

The purpose of this problem is to practice some R skills and review a critical topic in applied mathematics: Taylor series.

a\. Find the degree 5 Taylor polynomial $P(x)$ centered at $x = 0$ for $f(x) = \cos x$. This is a paper-and-pencil calculation, but you should write up the solution here in your markdown document.

b\. Find an upper bound for the error in approximating $f(x)$ by $P(x)$ on the interval $[-\pi/4,\pi/4]$. This is also a paper-and-pencil calculation that you should write up here.

c\. Make a single plot with 3 different curves: $cos(x)$ and the 2nd and 4th degree Taylor approximations of $f(x)$ around $x=0$. Plot these on the interval $[-\pi/2,\pi/2]$.

### Problem 2 Solution

a\. Your solution goes here.

b\. Your solution goes here.

c\. Your solution goes here.

### Problem 3

Consider the polynomial $f(x) = x^5 + x^4 - 2x^3 + 4x^2 - 8x + 5$. For a vector of 5 million equally spaced values going from x = -100 to x = 100, evaluate the polynomial four different ways, time each one using commands from the `tictoc` package, and compare. You should make your vector of x values using the `seq` command.

* Regular polynomial evaluation using a loop to iterate over each value in the vector.
* Regular polynomial evaluation using the entire vector at once.
* Horner's method using a loop to iterate over each value in the vector.
* Horner's method using the entire vector at once.

In addition to stating the times for each calculation, identify the slowest, the fastest, and the ratio of slowest to fastest.

### Problem 3 Solution

Your solution goes here.

### Problem 4

You read three articles on applications of linear algebra:

* [The Mathematics of Mass Testing for COVID-19](https://sinews.siam.org/Details-Page/the-mathematics-of-mass-testing-for-covid-19)
* [Machine Learning Has Been Used to Automatically Translate Long-Lost Languages](https://www.technologyreview.com/2019/07/01/65601/machine-learning-has-been-used-to-automatically-translate-long-lost-languages/)
* [The Best Bits](https://www.americanscientist.org/article/the-best-bits)

Choose one of these articles and write a brief paragraph about it in which you explain the role that linear algebra plays. Strive to be specific. For instance, rather than writing "linear algebra is used to make pooled testing more efficient," try to explain the actual mathematical ideas used.

### Problem 4

Your solution goes here.